Measuring the double-drift illusion with hand tracking: capturing illusion resets

### 2020-09-21 Manuscript.Rmd
incorporated a lot of Patrick's feedback

If a gabor pattern drifts in one direction while its internal texture drifts in the orthogonal direction, its perceived direction deviates strongly from its true direction and is instead some combination of its real external motion and its internal motion [@Tse2006]. The perceived location of such a double-drift stimulus periodically resets toward the veridical position. There may indeed be a simple perdiodicity underlying these resets. However, since both the veridical and perceived location are available to the brain, these resets may also occur when these two positions are too far from each other to come from the same source stimulus. Here we test which of the two explanations is more likely. In a first experiment, we had participants continuously track the position of the perceived position of the stimulus, using a stylus on a drawing tablet. While the orientation of the recorded trajectories reflect stimulus properties, there were no spontaneous resets. That’s why, in a second experiment, we had participants re-trace the perceived trajectory of a short stimulus presentation, or indicate the initial direction of the stimulus they perceived. We found resets here, and they indicate a set spatial limit to the distance between the veridical and perceived location of the double-drift stimulus.

@@

Encoding the position of objects in the world is necessary for the myriad of visually guided motor tasks that we do every day. It is no surprise then that the brain has several mechanisms to gauge and update the position of objects in the world. For example, when a target is moving, its motion can influence it perceived location, making it appear to be slightly ahead of its true location, e.g. the flash lag [@Nijhawan1994]; or the flash grab effect [@Cavanagh2013]. These cases of position extrapolation may be functional, helping the motor system overcome inevitable neural delays in targeting a moving object [@Nijhawan1994; @Duhamel1992]. A stronger and very different motion-induced position shift arises when a moving gabor with internal motion orthogonal to its path is viewed in the periphery [@Tse2006; @Shapiro2010; @Gurnsey2012; @Kwon2015; @Lisi2015]. This double-drift stimulus (Fig. 1, top left) generates extreme misjudgments of the moving target’s location which may deviate by as much as several degrees of visual angle from its true location. Surprisingly, this extraordinary perceptual illusion does not affect eye movements to the gabor: saccades to these targets are determined by their physical not their perceived location [@Lisi2015]. This result indicates that both true and illusory locations are available in the brain. Indeed, although immediate eye-movements go towards the true position, delayed eye movements and pointing go to the perceived position [@Lisi2017; @Massendari2018].

@@

The strength of the double-drift illusion probably arises because of the very poor location information for a gabor pattern in the periphery when it has the same mean luminance as the background. With poor positional certainty, the motion information contributes to the location estimate, generating a shift in position. If the gabor itself is not moving [@DeValois1991], the shift saturates after about 100 ms [@Chung2007], perhaps because the accumulating evidence of actual location of the stationary gabor patch is sufficient to anchor its position. However, when the gabor itself also moves as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues away from its physical location up to several times the size of the gabor. At some point, the illusion resets, the perceived location jumps back to its physical location and then accumulation begins again (Fig. 1, bottom left) or it saturates. This long duration before resetting or saturation is well beyond the integration period of other visual effects and into the range of short-term memory. Given that both true and illusory positions are available, it could be that instead of some integration period, it is the interplay between the two position signals that leads to resets.

Here, we examine the double-drift illusion by asking participants to trace out its path with a stylus on a tablet that is optically aligned with the monitor surface but with the hand not visible. Earlier results for pointing to the double-drift target [@Lisi2017] show that the illusion persists for manual pointing. The ability to manually track the perceived position will also allow us to study the resets – the adjustment of perceived location when the illusion has deviated too much or too long from the physical location. It could be that the imprecise, but true, retinal position signal limits the illusion, or it could be that – like other illusions, such as bi-stable stimuli – there is a simple periodicity underlying changes in perception. Here we capture where and when observers experience resets by having them draw the perceived trajectory of the stimulus while seeing it (a tracking procedure), or after having seen it (a retrace procedure). By investigating the time and location of resets in the recorded trajectories, we can distinguish between a spatial limit to the illusion or a temporal limit.

@@

For the first task in experiment 1, 4 participants were recruited from the lab (3 female; ages 21 – 31, mean: 25.75). For the second task in experiment 1, 6 participants were recruited from the lab (4 female; ages 19 – 26, mean: 21.6). For experiment 2, 9 participants were recruited from the lab (6 female; ages 19 - 27, mean: 22.8). All participants reported to be right handed and have either normal or corrected to normal vision. Procedures were approved by York’s Human Participants Review Committee, and all participants provided prior, written, informed consent.

@@

Participants were to be made For reporting degrees of visual angle that stimuli spanned ( _dva_ ), we assume the participant is close to the setup, with their eyes at the height of the monitor. Since there was no requirement to use the forehead rest, the latter is probably only true for the tallest participants, so that reported _dva's_ are underestimating the size of stimuli.

@@

In this experiment, two tasks were performed by two separate groups of participants. Both tasks asked participants to use a stylus on a drawing tablet to track a moving gabor with some amount of internal motion that elicited the double-drift illusion. The idea was that whenever spontaneous resets ofthe illusion occur, these would be recorded in the tracking data as sudden changes in heading (cf. Fig 1, bottom left, the dashed purple lines).

@@

In Figure 2 (top), we plot the average instantaneous heading during the middle second of each 2 _s_ pass (as tracking was noisy at the inversion points) for each of the 5 internal motion speeds (normalized for the external motion). The average angle of the tracked path is given in degrees deviation from straight forward. It seems that continuous tracking is sensitive to the strength of the illusion, and this is confirmed by a repeated-measures F-test on the average angles, using internal motion as a within-subjects factor (_F_(4,12)=102.6, _p_<.001, $\eta^2$=0.97). With FDR-corrected, paired t-tests, we find a difference in heading between 0 _cps_ internal motion and both 1 _cps_ ( _p_=.022) and -1 _cps_ ( _p_=.025) as well as between -3 and -1 _cps_ and 3 and 1 _cps_ (both _p_ <.001). This means we can use the averaged angle for each of the internal motion conditions as a measure of the strength of the illusion.

@@

In order to detect potential resets, we consider that in the 0 internal motion condition there is no illusion, so that we measure the tracking of the veridical path of the stimulus. Since both strong, "jump" resets and weak "hit-the-wall" resets (see Fig 1, bottom left) would change percieved direction back to or crossing the veridical motion path, their smooth-spline interpolated paths would contain samples similar in heading to the 0 internal motion condition. We use a 95% confidence interval of heading in the 0 internal motion condition to select trajectories from conditions with non-zero heading that may show a reset. These trajectories are drawn in red (only 1 per condition/participant), but as can be seen, the deviations from the condition's dominant heading occur just before or after the external motion reversals, so they may simply reflect difficulty of tracking the stimulus, rather than resets.

@@

Here we used a re-tracing procedure where participants drew the perceived path of a double-drift stimulus making a single pass across the workspace. This provided more time for the participants to reproduce any resets, i.e. without the need for real-time tracking. Nakayama & Holcombe [-@Nakayama2019; -@Nakayama2019b] have also used a re-tracing paradigm, in their case to study triggered resets. The re-tracing task was done in half of 8 blocks. In the other half, participants oriented a line to indicate the initial perceived direction of the stimulus motion. Block types were alternated, and the order was counterbalanced across participants. Each block used the 6 combinations of two external speeds ( corresponding to 3 or 4 _s_ presentation time ), and three internal speeds ( 2, 3 or 4 _cps_ ), 6 times. Within each of the two tasks, each of the combinations of internal and external speed was presented 24 times in total.

@@

## Comparison with Sirui's data set

While here we find that a spatial limit is the better explanation for resets, other evidence may point at a temporal limit instead. In an as of yet unpublished study [Sirui] obtained the point of subjective equivalence (PSE) of a double drift stimulus, and a static point that appeared some time after the double drift stimulus disappeared. The double-drift stimulus was on a 2 or 4 DVA length trajectory, and disappeared after 1, 2 or 3 seconds. The PSE's can be modelled very accurately with a model that assumes a fixed chance of a reset per unit of time (~.7%/s), that is: it explains the data using only a temporal bound on the illusion. Interestingly, this fully contradicts the current findings. The model predicts that the PSE decreases over time as it becomes more likely that a reset has occurred. We will compare the two data sets here.

```{r fig8, echo=FALSE, message=FALSE, fig.width=8, fig.height=3.5, fig.cap="\\label{fig:fig8}**Figure 8. PSE's of double-drift position and continuous tracking**"}
plotDeviations()
```

If the PSE's express the average perceived deviations from the actual trajectory of the double-drift stimulus, then they should show a similar pattern as the averaged trajectories in experiment 2. In order to be able to compare the two data sets, we will include all trajectories where our simple algorithm did not detect a reset, which should also be included in the PSE data. Further, we scale the trajectories, so that the Y end-point matches the duration of stimulus. Since the PSE data was collected in degrees visual angle, we convert the unit to degrees visual angles, using the estimate at the centre of the workspace. For the 4-second stimuli, we removed the last second. Figure 8 shows the individual participants' and the averaged trajectories (purple, red, orange) as well as the PSE's obtained in the other study (shown separately for 2 and 4 DVA stimulus trajectories) in gray.

For all internal motion speeds, we see a very similar average trajectory, that does not get closer to veridical over time. We verified this with a repeated measures ANOVA on perceived deviation from veridical stimulus position, using internal motion speed (2, 3, or 4 cycles/s) and time (1, 2, or 3 seconds after stimulus onset) as factors. If the trajectory data from experiment 2 here follows the same pattern as the PSE data, there should be an effect of time, and perhaps an interaction with internal motion speed. We find no effect of time ( _F_(2,16)=0.011, _p_=.957, $\eta^2$<.001), no effect of internal motion speed ( _F_(2,16)=0.127, _p_=.814, $\eta^2$=.0015) and no interaction ( _F_(4,32)=0.361, _p_=.635, $\eta^2$=.002). This means that we can't replicate the pattern found with the PSE approach. There may be several reasons for this. First, we notice that the trajectory data is highly dependent on participants. It could be that this method provides very noisy estimates, hiding any effect. It could also be that the effects truly vary a lot between participants. Since we can see that some participants do seem to follow the pattern found with the PSE approach, and some show a pattern of increasing average deviation between perceived and true position, it may that the difference between the data sets can simply be explained by random sampling of relatively small numbers of participants.

@@

Recently, it has been suggested that resets can be triggered by planned eye-movements to the stimulus [@Nakayama2019; @Nakayama2019b], particularly "jump" resets. In this experiment, we did not control for eye-movements with an eye-tracker, so we can't directly investigate this idea. Given a pre-motor theory of attention like mechanism, when spontaneously planned eye-movements trigger (jump) resets, such resets could occur at any point during the stimulus pass. However, it seems that for the participants that display "jump" resets in their re-tracing data, these happen at a fairly consistent location, which suggests that jump resets are instead stimulus triggered. This should however be tested with eye-tracking data, so that this will remain a speculation here.

>>>>

If a gabor pattern drifts in one direction while its internal texture drifts in the orthogonal direction, its perceived direction deviates strongly from its true direction and is instead some combination of its real external motion and its internal motion [@Tse2006]. The perceived location of such a double-drift stimulus periodically resets toward the veridical position. There may indeed be a simple perdiodicity underlying these resets. However, since both the veridical and perceived location are available to the brain, these resets may also occur when these two positions are too far from each other to come from the same source stimulus. Here we test which of the two explanations is more likely. In a first experiment, we had participants continuously track the position of the perceived position of the stimulus, using a stylus on a drawing tablet. While the orientation of the recorded trajectories reflect stimulus properties, there were no spontaneous resets. In a second experiment, we had participants re-trace the perceived trajectory of a short stimulus presentation, or indicate the initial direction of the stimulus they perceived. We found resets here, and they indicate a set spatial limit to the distance between the veridical and perceived location of the double-drift stimulus.

@@

Encoding the position of objects in the world is necessary for the myriad of visually guided motor tasks that we do every day. It is no surprise then that the brain has several mechanisms to gauge and update the position of objects in the world. For example, when a target is moving, its motion can influence its perceived location, making it appear to be slightly ahead of its true location, e.g. the flash lag [@Nijhawan1994]; or the flash grab effect [@Cavanagh2013]. These cases of position extrapolation may be functional, helping the motor system overcome inevitable neural delays in targeting a moving object [@Nijhawan1994; @Duhamel1992]. A stronger and very different motion-induced position shift arises when a moving gabor with internal motion orthogonal to its path is viewed in the periphery [@Tse2006; @Shapiro2010; @Gurnsey2012; @Kwon2015; @Lisi2015]. This double-drift stimulus (Fig. 1, top left) generates extreme misjudgments of the moving target’s location which may deviate by as much as several degrees of visual angle from its true location. Surprisingly, this extraordinary perceptual illusion does not affect eye movements to the gabor: saccades to these targets are determined by their physical not their perceived location [@Lisi2015]. And although immediate eye-movements go towards the true position, delayed eye movements and pointing go to the perceived position [@Lisi2017; @Massendari2018]. These findings suggest that both the true and illusory positions are available in the brain.

@@

The strength of the double-drift illusion probably arises because of the very poor location information for a gabor pattern in the periphery when it has the same mean luminance as the background. With poor positional certainty, the motion information contributes to the location estimate, generating a shift in perceived position. If the gabor itself is not moving [@DeValois1991], the shift saturates after about 100 ms [@Chung2007], perhaps because the accumulating evidence of actual location of the stationary gabor patch is sufficient to anchor its position. However, when the gabor itself also moves as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues away from its physical location up to several times the size of the gabor. At some point, the illusion resets, the perceived location jumps back to its physical location and then accumulation begins again (Fig. 1, bottom left) or it saturates. This long duration before resetting or saturation is well beyond the integration period of other visual effects and into the range of short-term memory. Given that both true and illusory positions are available, it could be that instead of some integration period, it is the interplay between the two position signals that leads to spontaneous resets. Alternatively, there is a simple periodicity or stochastics process underlies the changes in perception.

Here, we examine the double-drift illusion by asking participants to trace out its path with a stylus on a tablet that is optically aligned with the monitor surface but with the hand not visible. Earlier results for pointing to the double-drift target [@Lisi2017] show that the illusion persists for manual pointing. However, recent work has suggested that movement-related attention may lead to resets [@Nakayama2020], so that continuous real-time tracking of the gabor's position, may prevent the illusory shift in position. The ability to manually track the perceived position will also allow us to study the resets – the adjustment of perceived location when the illusion has deviated too much or too long from the physical location. Hence, we first test if the illusion occurs when people continunously track the position of the double drift stimulus. Then we try to capture where and when observers experience resets by having them draw the perceived trajectory of the stimulus while seeing it (a tracking procedure), or after having seen it (a retrace procedure). Investigating the time and location of resets in the recorded trajectories, allows us to distinguish between a spatial limit to the illusion or a temporal limit.

@@

For the first task in experiment 1, 4 participants were recruited from the lab (3 female; ages 21 – 31, mean: 25.75). For the second task in experiment 1, 6 participants were recruited from the lab (4 female; ages 19 – 26, mean: 21.6). For experiment 2, 9 participants were recruited from the lab (6 female; ages 19 - 27, mean: 22.8). All participants reported to be right handed and had either normal or corrected to normal vision. Procedures were approved by York’s Human Participants Review Committee, and all participants provided prior, written, informed consent.

@@

For reporting degrees of visual angle that stimuli spanned ( _dva_ ), we assumed the participants to be close to the setup, with their eyes at the height of the monitor. Since there was no requirement to use the forehead rest, the latter is probably only true for the tallest participants, so that reported _dva's_ are slightly underestimating the size of stimuli.

@@

In this experiment, two tasks were performed by two separate groups of participants. Both tasks asked participants to use a stylus on a drawing tablet to track a moving gabor with some amount of internal motion that elicited the double-drift illusion. The idea was to test both if tracking the illusion continuously would preserve the illusion and if the tracked paths reveal resets.

@@

In Figure 2 (top), we plot the average instantaneous heading during the middle second of each 2 _s_ pass (as tracking was noisy at the inversion points) for each of the 5 internal motion speeds (normalized for the external motion). The average angle of the tracked path is given in degrees deviation from straight forward. It seems that continuous tracking is sensitive to the strength of the illusion, and this is confirmed by a repeated-measures F-test on the average angles, using internal motion as a within-subjects factor ( _F_(4,12)=102.6, _p_<.001, $\eta^2$=0.97 ). With FDR-corrected, paired t-tests, we find a difference in heading between 0 _cps_ internal motion and both 1 _cps_ ( _p_=.022) and -1 _cps_ ( _p_=.025) as well as between -3 and -1 _cps_ and 3 and 1 _cps_ (both _p_ <.001). This makes it clear that continuous tracking preserves the illusion. We will use the averaged angle for each of the internal motion conditions as a measure of the strength of the illusion.

@@

In order to detect potential resets, we consider that in the 0 internal motion condition there is no illusion, so that we measure the tracking of the veridical path of the stimulus. Since both strong, "jump" resets and weak "hit-the-wall" resets (see Fig 1, bottom left) would change percieved direction back to or crossing the veridical motion path, their smooth-spline interpolated paths would contain samples similar in heading to the 0 internal motion condition. We use a 95% confidence interval of heading in the 0 internal motion condition to select trajectories from conditions with non-zero heading that may show a reset. These trajectories are drawn in red (only 1 per condition/participant), but as can be seen, the deviations from the condition's dominant heading occur just before or after the external motion reversals, so they may simply reflect difficulty of tracking the stimulus at the reversal points, rather than resets.

@@

Here we used a re-tracing procedure where participants drew the perceived path of a double-drift stimulus making a single pass across the workspace. This provided more time for the participants to reproduce any resets, i.e. without the need for real-time tracking. Nakayama & Holcombe [-@Nakayama2020] have also used a re-tracing paradigm, in their case to study triggered resets. The re-tracing task was done in half of 8 blocks. In the other half, participants oriented a line to indicate the initial perceived direction of the stimulus motion. Block types were alternated, and the order was counterbalanced across participants. Each block used the 6 combinations of two external speeds ( corresponding to 3 or 4 _s_ presentation time ), and three internal speeds ( 2, 3 or 4 _cps_ ), 6 times. Within each of the two tasks, each of the combinations of internal and external speed was presented 24 times in total.

@@

## Comparison with PSE data

While here we find that a spatial limit is the better explanation for resets, other evidence may point at a temporal limit instead. In an as of yet unpublished study [Sirui] obtained the point of subjective equivalence (PSE) of a double drift stimulus, and a static point that appeared some time after the double drift stimulus disappeared. The double-drift stimulus was on a 2 or 4 DVA length trajectory, traversed in 1, 2 or 3 seconds. The PSE's can be modelled very accurately with a model that assumes a fixed chance of a reset per unit of time (~0.7%/s), that is: it explains the data using only a temporal bound on the illusion. Interestingly, this fully contradicts the current findings. The model predicts that the PSE decreases over time as it becomes more likely that a reset has occurred. We will compare the two data sets here.

```{r fig8, echo=FALSE, message=FALSE, fig.width=6, fig.height=8, fig.cap="\\label{fig:fig8}**Figure 8. PSE's of double-drift position and continuous tracking**"}
plotAverageTrajectories()
```

If the PSE's express the average perceived deviations from the actual trajectory of the double-drift stimulus, then they should show a similar pattern as the averaged trajectories in experiment 2. In order to be able to compare the two data sets, we will include all trajectories, including those where our simple algorithm did not detect a reset, as such trials should also be included in the PSE data. Further, we scale the trajectories on both axes, so that the Y end-point matches the duration of stimulus. Since the PSE data was collected in degrees visual angle, we convert the unit to degrees visual angles, using the estimate at the centre of the workspace. For the 4-second stimuli, we removed the last second. Figure 8 shows the individual participants' and the averaged trajectories (purple, red, orange) as well as the PSE's obtained in the other study (shown separately for 2 and 4 DVA stimulus trajectories) in gray.

For all internal motion speeds, we see a very similar average trajectory, that does not get closer to veridical over time. We verified this with a repeated measures ANOVA on perceived deviation from veridical stimulus position, using internal motion speed (2, 3, or 4 cycles/s) and time (1, 2, or 3 seconds after stimulus onset) as factors. If the trajectory data from experiment 2 here follows the same pattern as the PSE data, there should be an effect of time, and perhaps an interaction with internal motion speed. We find no effect of time ( _F_(2,16)=0.011, _p_=.957, $\eta^2$<.001), no effect of internal motion speed ( _F_(2,16)=0.127, _p_=.814, $\eta^2$=.0015) and no interaction ( _F_(4,32)=0.361, _p_=.635, $\eta^2$=.002). This means that we can't replicate the pattern found with the PSE approach. There may be several reasons for this. First, we notice that the trajectory data is highly dependent on participants. It could be that the re-tracing method provides very noisy estimates, hiding any effect. It could also be that the effects truly vary a lot between participants. Since we can see that some participants do seem to follow the pattern found with the PSE approach, and some show a pattern of increasing average deviation between perceived and true position, it may that the differences between the data sets can simply be explained by random sampling of relatively small numbers of participants.

@@

Recently, it has been suggested that resets can be triggered by planned eye-movements to the stimulus [@Nakayama2020], particularly "jump" resets. In this experiment, we did not control for eye-movements with an eye-tracker, so we can't directly investigate this idea. Given a pre-motor theory of attention like mechanism, when spontaneously planned eye-movements trigger (jump) resets, such resets could occur at any point during the stimulus pass. However, it seems that for the participants that display "jump" resets in their re-tracing data, these happen at a fairly consistent location, which suggests that jump resets are instead stimulus triggered. This should however be tested with eye-tracking data, so that this will remain a speculation here.
